A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2016-10-11 doc/paper1-submission/paper4.tex
Update.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}
\newcommand{\otl}{OTLv4}
\newcommand{\otlprop}{OTLv4$^\prime$}

@@

\subsubsection{Annotation}

@@

To reveal the connections between the groupings found in the a summary
supertree and the input trees, propinquity uses a few Python scripts
and the \texttt{otc-annotate-synth} tool from otcetera to create an
annotations file describing the pipeline used and the connections
between phylogenetic information in the inputs and the summary. The
JSON file produced by \texttt{otc-annotate-synth} encodes a ``nodes''
property that holds a mapping between a node name for the summary
tree (using the naming convention described in the previous section)
and a node provenance object categorizes the relationship between
the node and the inputs. The node provenance object for node $x$
uses several properties to categorize the relationship between the
node and the inputs; each property in the node provenance object maps
to a structure storing the tree ID and node IDs for the input tree
nodes.

Conceptually, this annotation operation in equivalent to considering
every node $j$ in each input tree $i$ and the summary tree node
$x$. Because the vast majority of nodes in the input studies will
be compatible but not directly relevant to node $x$ we do not list
all of the compatible groupings. If node $x$ is not included in the
induced tree $\mathcal{\mathbb{S}}(i)$, then none of the nodes of
tree $i$ will be referred to in the annotations for node $x$. Even
if $x$ is included in $\mathcal{\mathbb{S}}(i)$, many of the nodes
of $T_{i}$ will be compatible with $x$ while being relevant to other
parts of the summary tree. The only input nodes listed for node $x$
are with rooted taxon bipartitions which conflict with, are displayed
by, or are resolved by the the rooted taxonomic bipartition associated
with node $x$. All input nodes that cannot be displayed by any supertree
that contains $x$ are stored in a ``conflicts\_with'' property
of the node provenance object. If node $j$ of $T_{i}$ is displayed
by the summary tree and $x$ is part of the path of $\mathcal{\mathbb{S}}(i)$
that displays the split between descendants of $j$ and other taxa,
then a reference to the node $j$ will be in the node provenance object.
The exact categorization of this annotation will depend on the configuration
of node $x$ on the induced tree $\mathcal{\mathbb{S}}(i)$:
\begin{itemize}
\item if $x$ is a terminal path in the induced tree then node $j$ will
be listed in the ``terminal'' property;
\item if $x$ is along an internal path that contains some nodes with out-degree
equal to 1, then node $j$ will be listed in the ``partial\_path\_of''
property; and
\item if $x$ is along an internal path without any node of out-degree 1,
then node $j$ will be listed in the ``supported\_by'' property,
because node $j$ supports the existence of grouping $x$ in the sense
that collapsing the edge that separates $x$ from its parent would
cause the summary tree to no longer display node $j$.
\end{itemize}
These three relationships are illustrated in Figure \ref{fig:displayed-annotations}.
Finally, if $T_{i}$ does not display $x$ from $\mathcal{\mathbb{S}}(i)$,
but there exists an unresolved node $j$ in $T_{i}$ which could be
resolved such that the tree would then display $x$, then a reference
to node $j$ will be listed in the ``resolves'' property of node
$x$.

The otcetera annotation tool can also detect cases in which including
information from node $j$ in $T_{i}$ could further resolve a polytomy
$x$ in the summary tree; such a case would be annotated using the
``resolved\_by'' property of $x$. However, because of our goal
of excluding unnecessary polytomies, none of the nodes in propinquity's
summary tree will use this annotation when they are annotated with
the set of input trees.

@@

\caption{Representation of input splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & supported\_by & partial\_path\_of & terminal & conflicts\_with & resolves & resolved\_by\tabularnewline
\hline 
\otl{} & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
\hline 
\otl{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
\hline 
\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
\hline 
\otlprop{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
\hline 
\end{tabular}}

@@

\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & supported\_by & partial\_path\_of & terminal & conflicts\_with & resolves & resolved\_by\tabularnewline
\hline 
\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 3676 & 4\tabularnewline
\hline 
\otl{} only & 296 & 0 & 0 & 19 & 17 & 4\tabularnewline
\hline 
\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 3,883 & 0\tabularnewline
\hline 
\otlprop{} only & 19 & 0 & 0 & 300 & 224 & 0\tabularnewline
\hline 
\end{tabular}}

@@

We performed an annotation of the \otl{} tree to compare the ability
of the \otl{} tree and the \otlprop{} to represent splits in the input trees.

We ran propinquity on the same input
trees used to construct version Open Tree version 4. Unlike the version
4 tree, the tree from propinquity contained no unsupported nodes or
unnecessary polytomies. The number of displayed input splits increased
from 35,518 (for version 4) to 39,639; an 11.2\% increase. The number
of input splits that do not conflict with the summary tree, but are
not incorporated dropped from 2718 to 0. The number of input splits
that conflict with the phylogeny dropped from 2760 to 1357, a decrease
of 1357, or 49.2\%. %

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}
\newcommand{\otl}{OTLv4}
\newcommand{\otlprop}{OTLv4$^\prime$}
\newcommand{\supportedby}{\texttt{supported\_by}}
\newcommand{\partialpathof}{\texttt{partial\_path\_of}}
\newcommand{\terminal}{\texttt{terminal}}
\newcommand{\conflictswith}{\texttt{conflicts\_with}}
\newcommand{\resolves}{\texttt{resolves}}
\newcommand{\resolvedby}{\texttt{resolved\_by}}

@@

\subsubsection{Annotation}\label{sec:Annotation}

@@

To reveal the connections between the groupings found in the a summary
supertree and the input trees, propinquity uses a few Python scripts
and the \texttt{otc-annotate-synth} tool from otcetera to create an
annotations file describing the pipeline used and the connections
between phylogenetic information in the inputs and the summary.  The
JSON file produced by \texttt{otc-annotate-synth} encodes a ``nodes''
property that holds a mapping between a node name for the summary
tree (using the naming convention described in the previous section)
and a node provenance object that categorizes the relationship between
the node and the inputs. The node provenance object for node $x$
uses several properties to categorize the relationship between the
node and the inputs; each property in the node provenance object maps
to a structure storing the tree ID and node IDs for the input tree
nodes.

Conceptually, this annotation operation is equivalent to considering
every node $j$ in each input tree $i$ and the summary tree node
$x$. Because the vast majority of nodes in the input studies will
be compatible but not directly relevant to node $x$ we do not list
all of the compatible groupings. If node $x$ is not included in the
induced tree $\mathcal{\mathbb{S}}(i)$, then none of the nodes of
tree $i$ will be referred to in the annotations for node $x$. Even
if $x$ is included in $\mathcal{\mathbb{S}}(i)$, many of the nodes
of $T_{i}$ will be compatible with $x$ while being relevant to other
parts of the summary tree. The only input nodes listed for node $x$
are with rooted taxon bipartitions which conflict with, are displayed
by, or are resolved by the the rooted taxonomic bipartition associated
with node $x$. All input nodes that cannot be displayed by any supertree
that contains $x$ are stored in a ``\conflictswith{}'' property
of the node provenance object. If node $j$ of $T_{i}$ is displayed
by the summary tree and $x$ is part of the path of $\mathcal{\mathbb{S}}(i)$
that displays the split between descendants of $j$ and other taxa,
then a reference to the node $j$ will be in the node provenance object.
The exact categorization of this annotation will depend on the configuration
of node $x$ on the induced tree $\mathcal{\mathbb{S}}(i)$:
\begin{itemize}
\item if $x$ is on a terminal path in the induced tree then node $j$ will
be listed in the ``\terminal{}'' property;
\item if $x$ is along an internal path that contains some nodes with out-degree
equal to 1, then node $j$ will be listed in the ``\partialpathof{}''
property; and
\item if $x$ is along an internal path without any node of out-degree 1,
then node $j$ will be listed in the ``\supportedby{}'' property,
because node $j$ supports the existence of grouping $x$ in the sense
that collapsing the edge that separates $x$ from its parent would
cause the summary tree to no longer display node $j$.
\end{itemize}
These three relationships are illustrated in Figure \ref{fig:displayed-annotations}.
Finally, if $T_{i}$ does not display $x$ from $\mathcal{\mathbb{S}}(i)$,
but there exists an unresolved node $j$ in $T_{i}$ which could be
resolved such that the tree would then display $x$, then a reference
to node $j$ will be listed in the ``\resolves{}'' property of node
$x$.

The otcetera annotation tool can also detect cases in which including
information from node $j$ in $T_{i}$ could further resolve a polytomy
$x$ in the summary tree; such a case would be annotated using the
``\resolvedby{}'' property of $x$. However, because of our goal
of excluding unnecessary polytomies, none of the nodes in propinquity's
summary tree will use this annotation when they are annotated with
the set of input trees.

@@

\caption{Representation of input splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
\hline 
\otl{} & 34,595 & 923 & 45,385 & 2760 & 473 & 2718\tabularnewline
\hline 
\otl{} only & 745 & 54 & 0 & 2055 & 0 & 2718\tabularnewline
\hline 
\otlprop{} & 38,521 & 1,118 & 45,385 & 1,357 & 515 & 0\tabularnewline
\hline 
\otlprop{} only & 4,671 & 249 & 0 & 652 & 42 & 0\tabularnewline
\hline
\label{tab:representinputs}
\end{tabular}}

@@

\caption{Representation of taxonomy splits in the \otl{} tree and the \otlprop{} tree.}
\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
 & \supportedby{} & \partialpathof{} & \terminal{} & \conflictswith{} & \resolves{} & \resolvedby{}\tabularnewline
\hline 
\otl{} & 125,384 & 0 & 2,424,255 & 1998 & 3676 & 4\tabularnewline
\hline 
\otl{} only & 296 & 0 & 0 & 19 & 17 & 4\tabularnewline
\hline 
\otlprop{} & 125,107 & 0 & 2,424,255 & 2,279 & 3,883 & 0\tabularnewline
\hline 
\otlprop{} only & 19 & 0 & 0 & 300 & 224 & 0\tabularnewline
\hline 
\label{tab:representtaxonomy}
\end{tabular}}

@@

We performed an annotation of both the \otl{} tree and the \otlprop{}
tree as described in section \ref{sec:Annotation} to assess the
ability of our new supertree method to represent splits from input
phylogenies (Table \ref{tab:representinputs}).  This annotation
classifies input phylogeny splits according to how they relate to a
summmary tree, so that each input edge falls in one of \supportedby{},
\partialpathof{}, \terminal{}, \conflictswith{}, or \resolvedby{}.

The number of displayed input splits (supported\_by +
partial\_path\_of) increased from 35,518 (for version 4) to 39,639; an
11.2\% increase. The \otlprop{} tree displays 4920 input splits that
are not displayed by the \otl{} tree, but does not display 799 input
splits that are displayed by the \otl{} tree.

Unlike the version 4 tree, the tree from propinquity contained no
unsupported nodes or unnecessary polytomies. The number of input
splits that do not conflict with the summary tree, but are not
incorporated (resolved\_by) dropped from 2718 to 0.

The number of input splits that conflict with the summary
(conflicts\_with) dropped from 2760 to 1357, a decrease of 1357, or
49.2\%.
